### 2023-03-29 paper-thesis/chapters/01-introduction.tex
Finish future work section

@@

Furthermore, the thesis contains some discussions:
\begin{itemize}
	\item We discuss how well semantic code completion works for our language (chapter~\ref{chap:editor-services})
	\item We compare our implementation with an implementation of the same language in Haskell (chapter~\ref{ch:comp-haskell}) and LambdaPi (chapter~\ref{ch:comp-lambdapi})
	\item We discuss how Spoofax can be improved to better support implementing dependently typed languages (chapter~\ref{ch:ergonomics})
	\item We discuss related work (chapter~\ref{ch:relatedwork}) and future work (chapter~\ref{ch:conclusion})
\end{itemize}

>>>>

These contributions are important for two reasons. Firstly from the perspective of a Spoofax researcher, developing a language with a complex type system in Statix tests the boundaries of what Statix can do, and how Statix could be further improved to support these usecases better. On the other hand, from the perspective of a researcher in dependent type systems, if we can show that Statix provides a good way to quickly implement a dependently typed language, then Spoofax could help with rapid prototyping of dependently typed languages.

@@

Furthermore, the thesis contains some discussions about the contributions:
\begin{itemize}
	\item We discuss how well semantic code completion works for our language (chapter~\ref{chap:editor-services})
	\item We compare our implementation with an implementation of the same language in Haskell (chapter~\ref{ch:comp-haskell}) and LambdaPi (chapter~\ref{ch:comp-lambdapi})
	\item We discuss how Spoofax can be improved to better support implementing dependently typed languages (chapter~\ref{ch:ergonomics})
	\item We discuss related work (chapter~\ref{ch:relatedwork}) and future work (chapter~\ref{ch:conclusion})
\end{itemize}

### 2023-03-29 paper-thesis/chapters/99-conclusion.tex
Finish future work section

\begin{itemize}
	\item Support for universe polymorphism~\cite[Universe Levels]{agda}, which allow functions and inductive data types to be polymorphic over a certain universe level. For example, this allows to define a \verb|List| data type that can be both a list of values and a list of types without defining it twice. We see no reason why this couldn't be cleanly implemented. 
	\item Another feature of Agda is implementing dependent pattern matching, as described in "Eliminating Pattern Matching without K"~\cite{patternmatching}. This is requires a desugaring stage, in which matches are desugared to eliminators, which should be implemented in Stratego before type checking. This should also be possible to implement, but it may be quite a bit of work.
	\item Inference in our language is currently implemented by explicitly specifying a placeholder \verb|_| where inference is required. In Agda we can specify an implicit argument, which is automatically inferred unless explicitly provided. We see no reason why this couldn't be cleanly implemented. 
	\item Instance Arguments are a special kind of implicit arguments, that are solved by the instance resolution algorithm of Agda~\cite[Instance Arguments]{agda}. They can be used similarly to type class constraints, in order to restrict the data types over which a term is polymorphic, for example requiring the terms to have an implementation of equality. The instance resolution algorithm should be possible to implementing by traversing the scope graph (since it is linear), and checking whether each let binding is a possible solution similar to the Agda implementation. 
	\item Agda has holes, which allows the user of the language to build proofs interactively. One of these interactions is proof search, when triggered Agda will try to find a term that fits the expected type of a hole. The interactive part of this was discussed in chapter~\ref{ch:relatedwork}, this is not yet supported by Spoofax. After support for this is added, a proof search algorithm could be implemented in Stratego, such as the structurally recursive unification algorithm described by McBride~\cite{proofsearch}.
	\item Agda allows function arguments to be marked as irrelevant, meaning that the value will be erased at runtime. An even stronger version of this feature is implementing a quantitive type system, such as in Idris~2~\cite{idris2}, which combines linear and dependent types. Implementing this in Spoofax should be possible, but it may be quite a bit of work as it is a complex type system.
	\item Agda has mixfix operators, which allows to define custom operators such as \verb|a and b|. This requires parsing to happen in multiple stages, where the first stage finds the mixfix operator definitions and the second stage parses expressions using them. This results in quite a complex parser, implementing these multiple stages in Spoofax using SDF3 is not possible without fundamental changes to the way that Spoofax works. One solution would be using an \emph{adaptive parsing algorithm}, which allows modification of the grammar during parsing. More research is needed to explore the possibilities of solving this problem.
	\item Modules
	\item Recursion
\end{itemize}

>>>>

\begin{itemize}
	\item Support for universe polymorphism~\cite[Universe Levels]{agda}, which allow functions and inductive data types to be polymorphic over a certain universe level. For example, this allows to define a \verb|List| data type that can be both a list of values and a list of types without defining it twice. We see no reason why this couldn't be cleanly implemented in Statix. 
	\item Another feature of Agda is implementing dependent pattern matching, as described in "Eliminating Pattern Matching without K"~\cite{patternmatching}. This is requires a desugaring stage, in which matches are desugared to eliminators, which should be implemented in Stratego before type checking. This should also be possible to implement, but it may be quite a bit of work.
	\item Inference in our language is currently implemented by explicitly specifying a placeholder \verb|_| where inference is required. In Agda we can specify an implicit argument, which is automatically inferred unless explicitly provided. We see no reason why this couldn't be cleanly implemented in Statix. 
	\item Instance Arguments are a special kind of implicit arguments, that are solved by the instance resolution algorithm of Agda~\cite[Instance Arguments]{agda}. They can be used similarly to type class constraints, in order to restrict the data types over which a term is polymorphic, for example requiring the terms to have an implementation of equality. The instance resolution algorithm should be possible to implementing by traversing the scope graph (since it is linear), and checking whether each let binding is a possible solution similar to the Agda implementation. 
	\item Agda has holes, which allows the user of the language to build proofs interactively. One of these interactions is proof search, when triggered Agda will try to find a term that fits the expected type of a hole. The interactive part of this was discussed in chapter~\ref{ch:relatedwork}, this is not yet supported by Spoofax. After support for this is added, a proof search algorithm could be implemented in Stratego, such as the structurally recursive unification algorithm described by McBride~\cite{proofsearch}.
	\item Agda allows function arguments to be marked as irrelevant, meaning that the value will be erased at runtime. An even stronger version of this feature is implementing a quantitive type system, such as in Idris~2~\cite{idris2}, which combines linear and dependent types. Implementing this in Spoofax should be possible, but it may be quite a bit of work as it is a complex type system.
	\item Agda has mixfix operators, which allows to define custom operators such as \verb|a and b|. This requires parsing to happen in multiple stages, where the first stage finds the mixfix operator definitions and the second stage parses expressions using them. This results in quite a complex parser, implementing these multiple stages in Spoofax using SDF3 is not possible without fundamental changes to the way that Spoofax works. One solution would be using an \emph{adaptive parsing algorithm}, which allows modification of the grammar during parsing. More research is needed to explore the possibilities of solving this problem.
	\item Another thing that could be explored is creating a module system for a dependently typed language in Spoofax. At the time of writing, Agda has a module system which is quite powerful (including parameterized modules), but the implementation is quite naive by substituting the modules into the code when imported~\cite[Modules]{agda}. Spoofax's scope graphs allow to create a module system with ease, simply by adding extra \emph{import} edges between scopes. Future research is needed on if Agda's module system could be implemented using scope graphs.
	\item Currently the language in the paper does not allow for recursive functions. Agda allows for functions that are structurally recursive, and uses termination checking to enforce this property. We see no reason why this couldn't be cleanly implemented in Statix, similar to the implementation of positivity checking in chapter~\ref{ch:datatypes}.
\end{itemize}

\subsection*{Implementing other complex type systems in Statix}

In this paper we have shown that a dependent type system can be implemented in Statix, one of the goals being to show the power of Statix. To further show this goal, more different kinds of type systems could be implemented in Statix:

\begin{itemize}
	\item There has not yet been an implementation of a sub-structural type system in Spoofax. A sub-structural type system is a type system that places restrictions on how often variables can be used. Now that languages with sub-structural type systems, such as Rust, are becoming popular, we should see if the goal of Statix to cover a broad range of type systems, still holds true for these. 
	\item Effect systems are a formal system that describe the side effects of a program. These allow programmers to specify what side effects a function has, which can then be verified by a type checker. Researching whether an effect system could be cleanly specified in Statix may lead to more insights on the limits of Statix.
\end{itemize}
